HealpixNet: Efficient spherical Convolutional Neural Network with HEALPix sampling for cosmological applications

### 2018-08-30 scnn.tex
small edits to experiments

As discussed in \secref{architecture}, the choice of a network architecture depends on the data and task at hand.
For our problem, where we assume that each pixel carries the same quantity of information about the cosmological model, a rotation invariant FCN architecture should be best.
The chosen network is defined as
\begin{equation*}
	FCN = SM \circ ST_{\text{mean}} \circ GC_2 \circ L_5 \circ L_4 \circ L_3 \circ L_2 \circ L_1,
\end{equation*}
\begin{center}
with \hspace{0.5cm} \begin{tabular}{lll}
   $L_1$ &  $=$ &$ P_4 \circ \sigma \circ BN \circ GC_{16}$, \\
   $L_2$ &  $=$ &$ P_4 \circ \sigma \circ BN \circ GC_{32}$, \\
   $L_3,L_4,L_5$ & $=$ &$ P_4 \circ \sigma \circ BN \circ GC_{64}$, \\
\end{tabular}
\end{center}
where $GC_F$ indicates a graph convolutional layer with $F$ output feature maps, $P_4$ is a pooling layer that divides the number of pixels by 4, $\sigma$ is a ReLU, and the Chebyshev polynomials in \eqnref{graph_convolution_cheby} are of degree $K=5$.
The layers $L_1$ to $L_5$ build the statistical evidence to classify each pixel of a down-sampled sphere. The last GC layer acts as a linear classifier and outputs two (non-normalized) predictions. The AV layer averages these predictions. Probabilities are computed by the SM layer.

@@

While testing the FCN and CNN variants, we made the following empirical observations.
First, training was more stable (in the sense that the loss had a more monotone decrease) when using Chebyshev polynomials, as in \eqnref{graph_convolution_cheby}, rather than monomials, as in \eqnref{graph_convolution_monomial}.
Nevertheless, we could not observe a significant difference in accuracy after convergence.
Second, using $\ell_2$ regularization does not help either with performance or training of the models, mostly because the models are not over-fitting.
Third, we recommend initializing the Chebyshev coefficients with a centered Gaussian distribution with standard deviation $\sqrt{2/(F_{in} \times (K + 0.5))}$, where $K$ is the degree of the Chebyshev polynomials and $F_{in}$ the number of input channels. This standard deviation has the property to keep the energy of the signal more or less constant across layers. \todo{Nath: cite the paper you read about that?}
% Nati to Michael: The problem with the previous claim is that it depends on the non-linarity. Furthermore, does it make sense to make such a claim if we use batch norm?
Finally, we observe that scaling the Laplacian's eigenvalues between $[-a, a]$, where $0 < a \leq 1$, significantly helps in stabilizing the optimization. We use $a = 0.75$ in our experiments.

@@

\begin{figure}
	\centering
	\includegraphics[width=\linewidth]{section_filter_last}
	\\
	\includegraphics[width=\linewidth]{gnonomic_filter_last}
	\caption{Random selection of 24 learned filters from the fifth spherical convolutional layer $L_5$. Top: section. Bottom: gnomonic view.}
	\label{fig:learned_filter}
\end{figure}

A common visualization to introspect and try to understand how a CNN function is to look at the learned filters.
Since our construction leads to almost (up to sampling irregularities) spherical filters, we plot in \figref{learned_filter} both the radial profile and a gnomonic projection on a plane of a random selection of learned filters from the last layer of the network.
While those particular filters were obtained from the experiment with order $o=2$ and a relative noise level of $2$, all trained networks presented visually similar gnomonic and profile patterns.
% The gnomonic projection clearly exhibits the sampling issue of HEALPix.
% \TK{what issue? Commenting out for now}
Details of how the convolution kernels are plotted are described in \ref{sec:filter_visualization}.
While it is usually difficult to interpret the shape of the filters, especially given the type of data, we can notice that they often have a ``peak'' like structure.
An example of filter interpretation was demonstrated in \citep{Ribli2018learningfrom}.
% Nati: Note that filter are almost not evolving during the training process. It is not surprising as our architecture is likely to be overparametrized. If we want to have interpretable filters, we probably have to use a much smaller architecture.

>>>>

As discussed in \secref{architecture}, the choice of a network architecture depends on the data and task at hand.
For our problem, where we assume that each pixel carries the same quantity of information about the cosmological model, a rotation invariant FCN architecture should be best.
The selected network is defined as
\begin{equation*}
	FCN = \underbrace{SM \circ AV \circ GC_2}_{\text{classifier}} \circ \underbrace{L_5 \circ L_4 \circ L_3 \circ L_2 \circ L_1}_{\text{feature extractor}},
\end{equation*}
\begin{center}
with \hspace{0.5cm} \begin{tabular}{lll}
   $L_1$ &  $=$ &$ P_4 \circ \sigma \circ BN \circ GC_{16}$, \\
   $L_2$ &  $=$ &$ P_4 \circ \sigma \circ BN \circ GC_{32}$, \\
   $L_3,L_4,L_5$ & $=$ &$ P_4 \circ \sigma \circ BN \circ GC_{64}$, \\
\end{tabular}
\end{center}
where $GC_F$ indicates a graph convolutional layer with $F$ output feature maps, $P_4$ is a pooling layer that divides the number of pixels by 4, $\sigma$ is a ReLU, and the Chebyshev polynomials in \eqnref{graph_convolution_cheby} are of degree $K=5$.
The layers $L_1$ to $L_5$ build the statistical evidence to classify each pixel of a down-sampled sphere. The last GC layer acts as a linear classifier and outputs two (non-normalized) predictions. The AV layer averages these predictions. Probabilities are computed by the SM layer.

@@

In all our experiments, we saw no sign of over-fitting.
With more data, HealpixNet can probably do even better.
The fact that the CNN lags behind the FCN is another sign of lack of data.

While testing the FCN and CNN variants, we made the following empirical observations.
First, training was more stable (in the sense that the loss decreased more smoothly) when using Chebyshev polynomials, as in \eqnref{graph_convolution_cheby}, rather than monomials, as in \eqnref{graph_convolution_monomial}.
Nevertheless, we could not observe a significant difference in accuracy after convergence.
Second, using $\ell_2$ regularization does not help either with performance or training of the models, mostly because the models are not over-fitting.
Third, we recommend initializing the Chebyshev coefficients with a centered Gaussian distribution with standard deviation $\sqrt{2/(F_{in} \times (K + 0.5))}$, where $K$ is the degree of the Chebyshev polynomials and $F_{in}$ the number of input channels. This standard deviation has the property to keep the energy of the signal more or less constant across layers. \todo{Nath: cite the paper you read about that?}
% Nati to Michael: The problem with the previous claim is that it depends on the non-linarity. Furthermore, does it make sense to make such a claim if we use batch norm?
Finally, we observe that scaling the Laplacian's eigenvalues between $[-a, a]$, where $0 < a \leq 1$, significantly helps in stabilizing the optimization. We use $a = 0.75$ in our experiments.

@@

\begin{figure}
	\centering
	\includegraphics[width=\linewidth]{section_filter_last}
	\\
	\includegraphics[width=\linewidth]{gnonomic_filter_last}
	\caption{Random selection of 24 learned filters from the fifth spherical convolutional layer $L_5$. Top: section. Bottom: gnomonic projection.}
	\label{fig:learned_filter}
\end{figure}

A common visualization to introspect and try to understand how a CNN function is to look at the learned filters.
Since our construction leads to almost (up to sampling irregularities) spherical filters, we plot in \figref{learned_filter} both the radial profile and a gnomonic projection on a plane of a random selection of learned filters from the last layer of the network.
While those particular filters were obtained from the experiment with order $o=2$ and a relative noise level of $2$, all trained networks presented visually similar patterns.
% The gnomonic projection clearly exhibits the sampling issue of HEALPix.
% \TK{what issue? Commenting out for now}
Details of how the convolution kernels are plotted are described in \ref{sec:filter_visualization}.
While it is usually difficult to interpret the shape of the filters, especially given the type of data, we can notice that they often have a ``peak'' like structure.
An example of filter interpretation was demonstrated in \citep{Ribli2018learningfrom}.
% Nati: Note that filter are almost not evolving during the training process. It is not surprising as our architecture is likely to be overparametrized. If we want to have interpretable filters, we probably have to use a much smaller architecture.
